evidence of adaptation-independent time-warped spike timing noise correlations in cortical networks

### 2020-08-11 manuscript.tex
Gut

\section{Introduction}
The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained. Such variability leads to the common assumption of population codes based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
% \cite{gerstner1997neural}, which have capacity beyond the original spike count \cite{}. 
% Continuous firing rates (averaged over time, trial or population ) 
% These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}. 
Trial-to-trial spike count variability is in fact correlated across the population in early sensory cortices \cite{reyes2015high, lin2015nature} and was shown to be modulated by a two-dimensional continuous state \cite{lin2015nature}. It is possible that spike time variability is modulated across a population by the same or similar low-dimensional underlying state. This study aims to characterise the nature of trial-to-trial spike time correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state. This offers a novel prediction of the encoding capacity of state-dependent spike time encoding. Characterising spike timing codes is important to understanding a variety of dynamic, homeostatic and plastic processes, as well as properties of action potential patterns directly involved in cognition.

@@

Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimulus \cite{averbeck2006neural}. `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}). As discussed, previous studies have demonstrated that spike count noise correlations are modulated across a population  by low-dimensional continuous latent states \cite{lin2015nature}. 
% For example, a 2-dimensional continuous latent state modulates responses in V1 \cite{lin2015nature}.

Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}. It is important to consider, that if downstream neurons are also modulated by the same population wide continuous latent state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the continuous latent state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.

@@

% This paper aims to characterise the precise nature of spike timing noise correlations for single stimuli to uncover the potential encoding capacity of a state-dependent spike time code. 
% The stimulus-dependency of noise correlations is left to future studies.  
Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
Processes of adaptation, which we define as different to latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig. X). Adaptation must therefore be controlled for when characterising adaptation-independent noise correlations.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of spike time noise correlations for first-spike patterns, towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies. A previous study analysing the same dataset found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set found spike count noise correlations and could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns \cite{reyes2015high}. The new analysis uses time series analysis to control for neural adaptation over all stimulus trials. adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

adaptation-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster. This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.

@@

\subsection*{Example first spike pair response distribution}

@@

\todo{Clear up final paragraph}
\todo{Differences standard deviation}

For the purpose of understanding, this section presents an example distribution of first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker (over a 50ms window following cortical activation onset). Only responses on `conjunctive trials' (trials in which both units spiked for the single stimulus) are shown. The orange Gaussian is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e. spike time jitter). This assumption fails to capture the distribution structure of this and many other spike pair distributions (Supplementary Vid. 1). 
% Moreover, very few different stimuli could be encoded with high-discriminability.

@@

Instead, first spike pairs appear to come from one of two clusters (blue and green dots). Moveover, the green cluster appears correlated with non-unitary gradient, suggestive of some underlying continuous state, which covaries the first spike times of the two units from trial-to-trial. Moreover, the apparent non-unitary gradient of correlation suggests that the distribution of differences varies with respect to the value of this underlying state, such that the state-conditioned distribution of relative first spike differences would be more compact than the general case, thus offering a higher potential capacity for stimulus encoding if downstream neurons could respond in a state-dependent manner.

>>>>

\section{Introduction}
The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained. Such variability leads to the common assumption of population codes based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
% \cite{gerstner1997neural}, which have capacity beyond the original spike count \cite{}. 
% Continuous firing rates (averaged over time, trial or population ) 
% These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}. 
In a recent study in early visual cortex, trial-to-trial spike count variability was shown to be correlated across the population and modulated by an underlying low 2-dimensional continuous latent state \cite{lin2015nature}. Other studies have reported only weak spike count correlations in sensory cortices \cite{reyes2015high}.

It is possible that spike time variability is caused and modulated across a population by a low-dimensional underlying state. This study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state. This offers a novel prediction of the encoding capacity of state-dependent spike time encoding. These results control for the occurrence of neural adaptation. The demonstrated nature of spike time noise correlations demonstrated offers an opportunity to  explain the report of only weak noise correlations in the previous characterisations of this data set \cite{reyes2015high}. Characterising spike timing codes is important to understanding a variety of dynamic, homeostatic and plastic processes, as well as properties of action potential patterns directly involved in cognition.

@@

Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimulus \cite{averbeck2006neural}. `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}). Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}. For fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.

% As discussed, previous studies have demonstrated that spike count noise correlations are modulated across a population  by low-dimensional continuous latent states \cite{lin2015nature}. 
% For example, a 2-dimensional continuous latent state modulates responses in V1 \cite{lin2015nature}.

It is important to consider, that if downstream neurons are also modulated by the same population wide continuous latent state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the continuous latent state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.

@@

% This paper aims to characterise the precise nature of spike timing noise correlations for single stimuli to uncover the potential encoding capacity of a state-dependent spike time code. 
% The stimulus-dependency of noise correlations is left to future studies.

Regardless of whether decoding neurons in the brain respond in a fixed-state or state-dependent manner, characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
Processes of adaptation, which we define as different to latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig. X). Adaptation must therefore be controlled for when characterising adaptation-independent noise correlations.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of spike time noise correlations for first-spike patterns, towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies.

A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}. The complex nature of spike time correlations uncovered in this study for the same data set may therefore offer an improved opportunity to study population-wide modulation by low-dimensional latent states. The new analysis uses time series analysis to control for neural adaptation over all stimulus trials. Adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

Adaptation-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster. This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.

@@

\subsection*{First spike pair response distributions}

@@

Analysing the joint responses of two or more units to repeated presentations of a single stimulus offers the opportunity to discern noise correlations caused by adaptative or non-adaptative processes. For the purpose of understanding, this section presents an example distribution of first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker (over a 50ms window following cortical activation onset). Only responses on `conjunctive trials' (trials in which both units spiked for the single stimulus) are shown.

First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus. Moreover, the green cluster appears correlated with non-unitary gradient, suggestive that some underlying processes (adaptative or non-adaptative) may covary the first spike times of the two units. Such correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid. 1).

This paper aims to quantify the occurrence of such correlated clusters and determine when adaptative and/or non-adaptative processes underlie such noise correlations. Importantly, the apparent non-unitary gradient of correlation of the green cluster in Fig. 3a suggests that the distribution of differences varies with respect to the possible adaptative and/or non-adaptative processes underlying the correlation.

The orange Gaussian in Fig. 3a is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e. spike time jitter). This assumption fails to capture the distribution structure of this and many other spike pair distributions (Supplementary Vid. 1). 
% Moreover, very few different stimuli could be encoded with high-discriminability.

@@

Moreover, the apparent non-unitary gradient of correlation suggests that the distribution of differences varies with respect to the value of this underlying state, such that the state-conditioned distribution of relative first spike differences would be more compact than the general case, thus offering a higher potential capacity for stimulus encoding if downstream neurons could respond in a state-dependent manner.
